---
title: "Mapping local values of measures"
bibliography: dbmss.bib
output: 
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
vignette: >
  %\VignetteIndexEntry{Bandwidth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include=FALSE}
# Installation of packages if necessary
InstallPackages <- function(Packages) {
  InstallPackage <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {
      install.packages(Package, repos="https://cran.rstudio.com/")
    }
  }
  invisible(sapply(Packages, InstallPackage))
}
# Add necessary packages here
Packages <- c("dplyr", "dbmss")
# Install them
InstallPackages(Packages)
knitr::opts_chunk$set(
  cache = TRUE, 
  message = FALSE, 
  warning = FALSE
)
set.seed(97310)
```


Local values of distance-based measures of spatial concentration can be mapped (@Marcon2023).


# Dataset simulation

We build a point pattern made of cases (the points of interest) and controls (the background distribution of points).

Cases are a Matérn [@Matern1960] point pattern with $\kappa$ (expected) clusters of $\mu$ (expected) points in a circle of radius *scale*.
Controls are a Poisson point pattern whose density $\lambda$ decreases exponentially along the y-axis (we will call "north" the higher y values).

```{r}
library("dplyr")
library("dbmss")
# Simulation of cases (clusters)
rMatClust(kappa = 10, scale = 0.05, mu = 10) %>% 
  as.wmppp -> 
  CASES
marks(CASES)$PointType <- "Cases"
# Number of points
CASES$n

# Simulation of controls (random distribution)
rpoispp(function(x, y) {1000 * exp(-2 * y)}) %>% 
  as.wmppp ->
  CONTROLS
marks(CONTROLS)$PointType <-"Controls"
# Number of points
CONTROLS$n

# Mixed patterns (cases and controls)
ALL <- superimpose(CASES, CONTROLS)
autoplot(ALL)
```


## Calculate and plot *M* Cases

```{r}
# Fix the number of simulations and the level of risk
NumberOfSimulations <- 1000
Alpha <- .01

# Calculate and plot M Cases
ALL %>% 
  MEnvelope(
    ReferenceType="Cases", 
    SimulationType = "RandomLocation",
    NumberOfSimulations = NumberOfSimulations,
    Alpha = Alpha,
    Global = TRUE
  ) ->
  M_env_cases

autoplot(M_env_cases)
```

The plot shows a clear relative concentration of cases.


## Map *M* results

To plot the individual values of *M* around each case, a distance must be chosen.
Then, the function must be computed at this distance with individual values.
Finally, a map is produced by smoothing the individual values and plotted.

```{r}
# Choose the distance to plot
Distance <- 0.1

# Calculate the M values to plot
ALL %>% 
  Mhat(
    r = c(0, Distance),
    ReferenceType = "Cases",
    NeighborType = "Cases", 
    # Individual must be TRUE
    Individual = TRUE
  ) -> 
  M_TheoEx

# Map resolution
resolution <- 512

# Create a map by smoothing the local values of M
M_TheoEx_map <- Smooth(
  # First argument is the point pattern
  ALL, 
  # fvind contains the individual values of M
  fvind = M_TheoEx, 
  # Distance selects the appropriate distance in fvind
  distance = Distance,
  Nbx = resolution, Nby = resolution
)

# Plot the point pattern with values of M(Distance)
plot(M_TheoEx_map, main = "")
# Add the cases to the map
points(
  ALL[marks(ALL)$PointType == "Cases"],
  pch = 20, col = "green"
)
# Add contour lines
contour(M_TheoEx_map, add = TRUE)
```

We can see that cases are concentrated almost everywhere (local *M* value above 1) because we chose a Matérn point pattern.

The areas with the higher relative concentration are located in the north of the map because the controls are less dense there.


## Compare with the density of cases

The density of cases is plotted.
High densities are *not* similar to high relative concentrations in this example because the control points are not homogeneously distributed.

```{r}
plot(density(CASES), main = "")
points(
  ALL[marks(ALL)$PointType == "Cases"],
  pch = 20, col = "green"
)
```


# References
