---
title: "Parallel simulations"
bibliography: dbmss.bib
output: 
  rmarkdown::html_document:
    toc: yes
    toc_float: yes
vignette: >
  %\VignetteIndexEntry{Parallel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, include=FALSE}
# Installation of packages if necessary
InstallPackages <- function(Packages) {
  InstallPackage <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {
      install.packages(Package, repos = "https://cran.rstudio.com/")
    }
  }
  invisible(sapply(Packages, InstallPackage))
}
# Add necessary packages here
Packages <- c("dplyr", "dbmss")
# Install them
InstallPackages(Packages)
knitr::opts_chunk$set(
  cache = TRUE, 
  message = FALSE, 
  warning = FALSE
)
set.seed(97310)
```


Simulations are necessary to compute the confidence envelopes of the main functions of the package.
They can be run in parallel thanks to the **future** and **doFuture** [@Bengtsson2021] packages.


# Example

Estimating the confidence envelope of the M function under the null hypothesis of random labeling of the point types relies on the `MEnvelope()` function.
Simulations are run sequentially.
The computing time can be measured.

```{r}
library("dbmss")
system.time(
  MEnvelope(
    paracou16, 
    r = NULL, 
    NumberOfSimulations = 5000, 
    Alpha = 0.05, 
    ReferenceType = "V. Americana", 
    NeighborType = "Q. Rosea", 
    SimulationType = "RandomLabeling", 
    Global = TRUE
  )  
)
```

It can be parallelized by setting a hardware strategy.
The following example launches several simultaneous copies of R on the local computer: it works on all operating systems.
The number of parallel processes is the number of logical CPUs minus 1.

```{r}
library("doFuture")
plan(multisession, workers = availableCores(omit = 1))
```

The same set of simulations is now run much faster with argument `parallel = TRUE`.

```{r}
system.time(
  MEnvelope(
    paracou16, 
    r = NULL, 
    NumberOfSimulations = 5000, 
    Alpha = 0.05, 
    ReferenceType = "V. Americana", 
    NeighborType = "Q. Rosea", 
    SimulationType = "RandomLabeling", 
    Global = TRUE, 
    parallel = TRUE
  )
)
```
The default strategy is `sequential`.
It should be restored.

```{r}
plan(sequential)
```

Other possible strategies are forking (`multicore`), unavailable on Windows, and `cluster`, on several computers.


See the documentation of the *future* package, namely `future::plan()`.

# Progress bar

Parallel simulations can report their progress thanks to the **progressr** package.
They ignore the `verbose` argument.
The progress bar must be activated and its interface chosen, for example by the following code, to be run before calling the `MEnvelope()` function:

```{r}
#| eval: false
library("progressr")
handlers(global = TRUE)
handlers("txtprogressbar")
```

Note that the choice of the progress bar is free: see `?progressr`.

# Limits

Setting up the parallel code and gathering the results takes time so the performance of sequential simulations may be better if they are fast.

  
# References
